---
title: "Weekly 2: Coffee Ratings"
author: "[Aiden Byrd]"
format:
  html:
    embed-resources: true
editor: visual
---

## Overview

You want to know which producers grow the best-rated coffee.

That sounds like a `group_by()` and a `summarise()`. The problem is that `producer` was typed by different people in different ways, so the same producer can appear under several labels, and some labels are not producers at all.

Your job is to work out how much of that you can fix, how much you shouldn't, and what that does to the answer.

A few things:

1.  Add `#| echo: false` to your code blocks.
2.  Use a new block when you move to a new task.
3.  Code goes in the blocks. **Answers go outside them**, as prose. A comment inside a code chunk is not an answer.
4.  Keep the original `producer` column. Every cleaned version goes in a **new** column.

Due Monday at 8pm.

## Your data

Import `coffee_ratings.csv` and inspect it.

```{r}
#| warning: false
#| echo: false
library(tidyverse)

coffee_ratings <- read_csv("coffee_ratings.csv")

glimpse(coffee_ratings)

```

## Before you start

> What does one row represent? One coffee with ratings, origin, year harvested, and more info on various types of coffee beans

> Is there a column that uniquely identifies a row? How would you check? No, we can check every column by comparing its number of distinct values to the total row count. If a column's distinct count matches the row count then every value is different and it could work as an ID, but none match so this data doesn't have a unique identifier

> Is `producer` a key? What does it mean that the same producer can appear on many rows? No it is not unique, the same producer showing up on multiple rows means that producer grew more than one coffee bean and got rated separately.

## How many producers?

Count the distinct values of `producer`.

```{r}
#| warning: false
#| echo: false
n_distinct(coffee_ratings$producer)

sum(is.na(coffee_ratings$producer))

n_distinct(coffee_ratings$producer, na.rm = TRUE)

```

> How many distinct values are there? 692 distinct values

> How many rows have no producer recorded? Does your distinct count include "missing" as one of its values? How would you check? 231 rows, yes that counts NA as a value, without the N/A its 691

> Why shouldn't you read that number as the number of distinct producers? Because the name of the producers can be typed differently, with different spacing, capitilization, or extra spaces

## Normalize the representation

Create a new column, `producer_clean`, that makes the producer labels consistent in case and whitespace. Keep `producer`.

```{r}
#| warning: false
#| echo: false
coffee_clean <- coffee_ratings |> 
  mutate(
    producer_clean = producer |> 
      str_to_upper() |> 
      str_squish()
  )

n_distinct(coffee_clean$producer_clean)
```

> How many distinct values does `producer_clean` have? 666

> Is that the number of producers now? What does the reduction establish, and what does it *not* establish? No 666 is just the number of distinct labels left after fixing spacing and correct casing, not a count of the real-world producers. Going from 692 to 666 shows that the data was formatting the labels differently. It does not establish the producer count because labels that differ in punctuation and different name have not been touched yet.

## Which labels actually merged?

A single number doesn't tell you what happened. Build a table showing every `producer_clean` value that more than one raw `producer` value was combined into, along with the raw values that went into it.

```{r}
#| warning: false
#| echo: false
coffee_clean |> 
  filter(!is.na(producer_clean)) |> 
  distinct(producer, producer_clean) |> 
  count(producer_clean, name = "n_raw_values") |> 
  filter(n_raw_values > 1) |> 
  left_join(
    coffee_clean |> distinct(producer, producer_clean),
    by = "producer_clean"
  ) |> 
  arrange(desc(n_raw_values), producer_clean)

```

> How many cleaned labels absorbed more than one raw label? 22

> Look through them. Is every merge one you would stand behind? A few of these merges are not producer names, like SEVERAL, VARIOS, VARIOUS.

> Is this table more useful than knowing the number of labels went down? Why? Yes this table lets us see what merged into what, without it we wouldn't know who our real producers are.

## Does it change anything?

Compare the La Plata coffees before and after normalizing. For each version, report the number of coffees and the mean and median `total_cup_points`.

```{r}
#| warning: false
#| echo: false
# BEFORE 
coffee_ratings |> 
  filter(producer == "La Plata") |> 
  summarise(
    n = n(),
    mean_score = mean(total_cup_points, na.rm = TRUE),
    median_score = median(total_cup_points, na.rm = TRUE)
  )

#AFTER 
coffee_clean |> 
  filter(producer_clean == "LA PLATA") |> 
  summarise(
    n = n(),
    mean_score = mean(total_cup_points, na.rm = TRUE),
    median_score = median(total_cup_points, na.rm = TRUE)
  )
```

> What changed more: the number of coffees or the average score? The number of coffees changed a lot more than the average score, The count went from 30 to 39, mean only moved from 83.65 to 83.51, and median moved from 83.75 to 83.58. Normalizing changed how many coffees we are counting for La Plata, but barely moved the actual rating numbers.

> Suppose you had reported how many coffees La Plata produced using the raw column. How far off would you have been? I would have used raw column and counted 30 coffees for La Plata, which was 9 coffees off, so close but not exactly.

## What case and whitespace didn't catch

Some labels still differ only in punctuation, i.e., a period, a comma, a slash.

Sort the distinct values of `producer_clean` and read through them. Near-duplicates end up next to each other.

```{r}
#| warning: false
#| echo: false

coffee_clean |> 
  distinct(producer_clean) |> 
  arrange(producer_clean) |> 
  print(n = 700)
```

> Find at least **three** pairs of labels that differ only in punctuation.
>
> AGRICOLA EL PILAR, S.A. vs AGRICOLA EL PILAR, SA.
>
> PACHUJ, S. A vs PACHUJ, S.A
>
> YEH, LI CHUAN vs YEH, LI CHUAN.

Now use exact replacement to make each pair consistent. Change only the values you identified and nothing else.

```{r}
#| warning: false
#| echo: false

coffee_clean <- coffee_clean |> 
  mutate(
    producer_clean = case_when(
      producer_clean == "AGRICOLA EL PILAR, SA." ~ "AGRICOLA EL PILAR, S.A.",
      producer_clean == "PACHUJ, S. A." ~ "PACHUJ, S.A",
      producer_clean == "YEH, LI CHUAN." ~ "YEH, LI CHUAN",
      TRUE ~ producer_clean
    )
  )

n_distinct(coffee_clean$producer_clean)


```

> For each change, say why you're confident it's the same producer."AGRICOLA EL PILAR, S.A." and "AGRICOLA EL PILAR, SA." are the same name with one missing period. "PACHUJ, S. A." and "PACHUJ, S.A" are different becasuse a space and a trailing period. "YEH, LI CHUAN" and "YEH, LI CHUAN." differ by a trailing period.

> How long did it take to find those by reading? How many labels did you have to look at? I made it so that I could see all 666 values, but I went down the list, copied the first word of the value, and control F that word and saw if there was common themes.

## The top producers

Using `producer_clean`, find the ten producers with the most coffees in the data.

```{r}
#| warning: false
#| echo: false

coffee_clean |> 
  filter(!is.na(producer_clean)) |> 
  count(producer_clean, sort = TRUE) |> 
  slice_head(n = 10)
```

Look at that list carefully.

> Is every entry in it a producer? If not, which ones aren't, and what do they mean? No. VARIOS is not a real producer, it is Spanish for various. In this data, it means a bunch of different unnamed producers and not one specific producer

> Is any single producer split across more than one row of the list?
>
> Yes, Ipanema shows up three different times

Now look at these four raw labels:

- Ipanema Agricola
- Ipanema Agricola S.A
- Ipanema Agrícola SA
- Ipanema Coffees

> **Should these be combined into one producer? Don't recode them. Explain whether combining them would be normalization or classification, and what evidence you'd need before doing it.**
>
> The first three in that list are the same name but formatted differently or have different punctuation, combining those would be normalization. I would combine them personally. Ipanema Coffees is a different name, and could be like the whole company, not just the Agricola producers. It would be classification and I personally would not combine those values.

> **If you did combine them, what would happen to your top-ten list?**
>
> If we combined these, the Ipanema count would be a lot higher in the top ten because we are combining likely one branch of Ipanema and the whole company, could also get repeated data.

## What this means for your question

> **A classmate says: "I reduced the producer column from 692 labels to 666, so the cleaned column is better." Is that justified? Explain.**
>
> Not justified, it just means some labels merged, you can´t tell from the numbers whether those merges were correct

> **You set out to find which producers grow the best-rated coffee. Given everything above, what would you need to tell someone before they trusted your answer?**
>
> I would tell them some rows have no producer or words that are not producers (Varios, Coop, etc), also Ipanema has four differently names rows, and we are not 100% sure if they are the same.

## Disclosure appendix

> What did you try first? Include the code you wrote and the actual error message, or a clear description of the problem if there was no eerror. I didn´t really have any major coding errors, had to look in the help section though for what goes inside mean and median functions since it´s been a while. I also kind of had a hard time starting out figuring out what pattern to use to find the similar producer names.
>
> If you used an AI system: the exact prompt, exactly what it returned, what it got wrong or left unclear, and what you changed.
>
> If you didn't use one, say so. I did not
